---
title: "Post Bariatric Surgery Depression Prevalence and Risk Factors"
author: "Osama Ahmed Abd El-Basset Ibrahim"
output: word_document
---

```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  dpi = 300)
```

 R version 4.5.1 (2025-06-13) was used to conduct the analysis. Descriptive statistics, including medians, inter-quartile ranges, and frequencies, were calculated to summarize the characteristics of the study population. A p-value of less than 0.05 was considered statistically significant. Additionally, logistic regression analysis was done to examine the impact of various factors on the likelihood of experiencing depression. 
 
```{r message=FALSE, warning=FALSE, include=FALSE}
library(readxl)
library(tidyverse)
library(forcats)
library(gtsummary)
library(broom.helpers)
library(patchwork)
library(rstatix)
library(flextable)


# importing the dataset 
post_bariatric_depression_data <- read_excel(path = "week 3 data.xlsx") 


# After first inspection, Abnormal height values were found 
# Doing further investigation 
post_bariatric_depression_data %>% filter(Height < 150)

# Changing the abnormal values 
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(Height = 
                                            ifelse(Height < 100, Height + 100, Height))

# Calculating BMI pre and post operation 
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(`BMI PreOp` = round(
  `Weight before surgery`/(Height/100)^2, digits = 1),
  `BMI PostOp` = round(
    `Current weight`/(Height/100)^2, digits = 1
  )) %>% select(-`Diagnosed with psychiatric illness?`, -`If yes, what?`)

# Expanding the comorbidities column into binary indicator variables for each condition
# And computing the total number of comorbidities for each patient
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  `Other Comorbidities` = as.integer(str_detect(Comorbidities, "others")),
  Hypertension = as.integer(str_detect(Comorbidities, "Hypertension")),
  Diabetes = as.integer(str_detect(Comorbidities, "Diabetes")),
  `High Cholesterol` = as.integer(str_detect(Comorbidities, "high cholesterol")),
  Asthma = as.integer(str_detect(Comorbidities, "asthma")),
  `Heart Disease` = as.integer(str_detect(Comorbidities, "heart disease")),
  `Number of Comorbidities` = as.numeric(`Other Comorbidities` + Hypertension + Diabetes + `High Cholesterol` + Asthma + 
    `Heart Disease`)
  # Reordering variables
) %>% select(-Comorbidities) %>% relocate(Hypertension, Diabetes, `Heart Disease`, Asthma, 
                                          `Other Comorbidities`,
                                          `Number of Comorbidities`, `High Cholesterol`,
                                          .after = `Educational Level`)

# Cleaning type of surgery column and segmenting it based on clinical knowledge to restrictive and malabsorptive surgereis 
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  Surgery_Type = case_when(
    `Type of surgery` %in% c("Sleeve gastrectomy", "Sleeve gastrectomy, Laparoscopic adjustable gastric banding",
                             "Laparoscopic adjustable gastric banding", "Sleeve gastrectomy, Balloon insertion",
                             "Balloon insertion") ~ "Restrictive",
    TRUE ~ "Malabsorptive or Mixed"
  ) 
) %>% relocate(Surgery_Type, .after = `Type of surgery`)

# Factoring and Reordering time since surgery variable
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  `Date since operation` = factor(`Date since operation`,levels = c("< 1 year", "1 - 2 years", "> 2 - 3 years",
                                             "> 3 years"))
)

# Factor coding the variables of the PHQ-9
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  across(c("Little interest or pleasure in doing things", "Feeling down, depressed, or hopeless",
           "Trouble falling or staying asleep, or sleeping too much",
           "Feeling tired or having little energy", "Poor appetite or overeating",
           "Feeling bad about yourself — or that you are a failure or have let yourself or your family down",
           "Trouble concentrating on things, such as reading the newspaper or watching television",
           "Moving or speaking so slowly that other people could have noticed? Or the opposite — being so fidgety or restless that you have been moving .around a lot more than usual",
           "Thoughts that you would be better off dead or of hurting yourself in some way"),
         ~ as.numeric(as.character(fct_recode(.x,
                                 "0" = "Not at all",
                                 "1" = "Several days",
                                 "2" = "More than half days",
                                 "3" = "Nearly everyday")))))

# Computing the PHQ-9 Score
post_bariatric_depression_data <- post_bariatric_depression_data %>% 
  mutate(`Score of PHQ-9` = `Little interest or pleasure in doing things` + 
           `Trouble falling or staying asleep, or sleeping too much` + 
           `Feeling tired or having little energy` + 
           `Feeling bad about yourself — or that you are a failure or have let yourself or your family down` + 
           `Trouble concentrating on things, such as reading the newspaper or watching television` + 
           `Moving or speaking so slowly that other people could have noticed? Or the opposite — being so fidgety or restless that you have been moving .around a lot more than usual` +
           `Thoughts that you would be better off dead or of hurting yourself in some way` +
           `Feeling down, depressed, or hopeless`)

# Categorize PHQ-9 total scores into standard depression severity levels
# based on Kroenke et al. (2001) scoring guidelines, then convert to
# an ordered factor for proper sorting in tables and plots
post_bariatric_depression_data <- post_bariatric_depression_data %>%
  mutate(`Depression Severity` = 
           case_when(
             `Score of PHQ-9` >= 0 & `Score of PHQ-9` <= 4 ~ "None-minimal",
             `Score of PHQ-9` >= 5 & `Score of PHQ-9` <= 9 ~ "Mild",
             `Score of PHQ-9` >= 10 & `Score of PHQ-9` <= 14 ~ "Moderate",
             `Score of PHQ-9` >= 15 & `Score of PHQ-9` <= 19 ~ "Moderately Severe",
             `Score of PHQ-9` >= 20 & `Score of PHQ-9` <= 27 ~ "Severe"
           )) %>% mutate(
             `Depression Severity` = factor(`Depression Severity`, levels = c(
               "None-minimal", "Mild", "Moderate", "Moderately Severe", "Severe"
             ))
           )



# Convert selected categorical columns to factors with title-cased labels
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  across(c("Gender", "Maraital Status", "Educational Level", "Postsurgical-complications",
           "visitng psychiatric PreOp", "visitng psychiatric PostOp"), ~ factor(str_to_title(.x)))
)

post_bariatric_depression_data <- post_bariatric_depression_data %>%
  mutate(`Educational Level` = fct_recode(`Educational Level`, 
                                          "Masters Or PhD" = "Masters Or Phd"))

# Reclassify "Illitrate" under "High School Or Less"
# to consolidate low-education categories into a single meaningful group
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  `Educational Level` = case_when(
    `Educational Level` %in% "Illetrate" ~ "High School Or Less", 
    TRUE ~ `Educational Level`
  )
)

# Re-apply factor with the correct ordered levels after recoding
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  `Educational Level` = factor(`Educational Level`, levels = c("High School Or Less" , "Diploma",
                                                           "Bachelor Degree","Masters Or PhD")))

post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(`Maraital Status` = 
                                            factor(`Maraital Status`, levels = c(
                                              "Single", "Married", "Divorced", "Widow"
                                            )))
  

# Rename columns to cleaner, more descriptive names for readability
# in output tables and figures
post_bariatric_depression_data <- post_bariatric_depression_data %>% rename("Visiting Psychiatric Pre-Operation" = `visitng psychiatric PreOp`,
                                          "Visiting Psychiatric Post-Operation" = `visitng psychiatric PostOp`,
                                          "Post Surgical Complications" = `Postsurgical-complications`) 

post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(`Post Surgical Complications` = 
                                            case_when(
                                              str_detect(`Post Surgical Complications`,"no|No") ~ "No",
                                              TRUE ~ `Post Surgical Complications`
                                            ))

# Calculate percent weight loss from pre-surgery weight to current weight
# Formula: ((Weight Before - Current Weight) / Weight Before) * 100
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  `Percentage of Weight Loss` = ((`Weight before surgery` - `Current weight`)/`Weight before surgery`)*100
)


# Apply the PHQ-9 diagnostic algorithm to flag depressive syndromes.
# Items scored >= 2 indicate symptoms present "more than half the days"
# or "nearly every day."
post_bariatric_depression_data <- post_bariatric_depression_data %>%
  mutate(
    # Count how many items are scored >= 2 (more than half the days or nearly everyday)
    items_high_frequency = rowSums(
      select(., 
             `Little interest or pleasure in doing things`,
             `Feeling down, depressed, or hopeless`,
             `Trouble falling or staying asleep, or sleeping too much`,
             `Feeling tired or having little energy`,
             `Poor appetite or overeating`,
             `Feeling bad about yourself — or that you are a failure or have let yourself or your family down`,
             `Trouble concentrating on things, such as reading the newspaper or watching television`,
             `Moving or speaking so slowly that other people could have noticed? Or the opposite — being so fidgety or restless that you have been moving .around a lot more than usual`,
             `Thoughts that you would be better off dead or of hurting yourself in some way`
      ) >= 2, na.rm = TRUE
    ),
    
    # Check if Item 1 OR Item 2 is >= 2
    core_symptoms_present = (
      `Little interest or pleasure in doing things` >= 2 | 
        `Feeling down, depressed, or hopeless` >= 2
    ),
    
    # MDD is suggested if BOTH conditions are met
    `Major Depressive Syndrome Suggested` = case_when(
      items_high_frequency >= 5 & core_symptoms_present == TRUE ~ "Yes",
      items_high_frequency < 5 | core_symptoms_present == FALSE ~ "No",
      TRUE ~ NA_character_
    ),
    `Other Depressive Syndrome Suggested` = case_when(
      items_high_frequency >= 2 & items_high_frequency <= 4 & core_symptoms_present == T ~ "Yes",
      TRUE ~ "No"
    ),
    `No Depression Suggested` = case_when(
      `Major Depressive Syndrome Suggested` == "No" & `Other Depressive Syndrome Suggested` == "No" ~ "Yes",
      TRUE ~ "No"
    )
  ) 


# Create a binary Depression variable: "Yes" if either depressive syndrome
# is suggested, "No" otherwise — used as the primary outcome variable
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  Depression = factor(case_when(
    `Major Depressive Syndrome Suggested` == "Yes" | `Other Depressive Syndrome Suggested` == "Yes" ~ "Yes",
    TRUE ~ "No"
  ))
) 


```
 
## Results
A total of 186 post-bariatric surgery patients were included in this study. The sample was predominantly female (66%) with a median age of 37 years (IQR: 29–44). The majority were married (61%) and held a bachelor's degree (51%), while 30% had a high school education or less. The median preoperative BMI was 46 kg/m² (IQR: 41–53), and restrictive procedures accounted for the vast majority of surgeries (84%). Regarding comorbidities, hypertension was the most prevalent condition (18%), followed by diabetes and high cholesterol (12% each), with 10% of patients reporting post-surgical complications and only 4.8% having visited a psychiatrist prior to surgery. 

#### Table 1: Sociodemographic profile of the study participants 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Summarize pre-operative and demographic variables for Table 1.
# Outputs a formatted flex table suitable for Word/HTML reports.
post_bariatric_depression_data %>% 
  select(Age, Gender, `Maraital Status`, `BMI PreOp`,`Educational Level`,
         Hypertension, Diabetes, `Heart Disease`, Asthma,
        `High Cholesterol`, `Other Comorbidities`,
        Surgery_Type, `Date since operation`, `Post Surgical Complications`,
        `Visiting Psychiatric Pre-Operation`) %>% 
  tbl_summary() %>% bold_labels() %>% as_flex_table()
```

Assessment of depressive sypmtoms using the PHQ-9 revealed a median total score of 5.0 (IQR: 2.0–9.0) across the sample. Fatigue and appetite disturbance were the most frequently endorsed symptoms, with 85% and 83% of patients reporting tiredness and appetite changes on at least several days, respectively. Sleep disturbance was also notable, with 22% of patients experiencing it nearly every day. In contrast, psychomotor symptoms and suicidal ideation were the least commonly reported, with 69% and 83% of patients denying psychomotor changes and thoughts of self-harm entirely, respectively — though 14% did endorse suicidal ideation at some frequency. Core depressive symptoms of anhedonia and depressed mood were reported on at least several days by 60% and 66% of patients, respectively.

### Figure 1: Distributionn of depression severity postbariatric surgery 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Quick exploratory bar chart of depression severity distribution.
# Percentages are calculated from counts and displayed above each bar.
post_bariatric_depression_data %>%
  count(`Depression Severity`) %>%
  mutate(
    Percentage = (n / sum(n)) * 100,
    `Depression Severity` = fct_reorder(`Depression Severity`, Percentage, .desc = TRUE)
  ) %>%
  ggplot(aes(x = `Depression Severity`, y = Percentage, fill = `Depression Severity`)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(
    aes(label = paste0(round(Percentage, 1), "%")),
    vjust = -0.5,
    fontface = "bold",
    size = 4
  ) +
  scale_fill_brewer(palette = "Reds", direction = 1) +
  scale_y_continuous(limits = c(0, 50)) +
  labs(
    x = "",
    y = "Percentage (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray50"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(face = "bold", size = 11, angle = 15)
  )
```
  
#### Table 2: Participants assessment of depression after bariatric surgery based on Patient Health Questionnaire-9 survey 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Summarize the distribution of responses for each PHQ-9 item
# along with the total PHQ-9 score; outputs a formatted flex table.
post_bariatric_depression_data <- post_bariatric_depression_data %>% mutate(
  across(c("Little interest or pleasure in doing things", "Feeling down, depressed, or hopeless",
           "Trouble falling or staying asleep, or sleeping too much",
           "Feeling tired or having little energy", "Poor appetite or overeating",
           "Feeling bad about yourself — or that you are a failure or have let yourself or your family down",
           "Trouble concentrating on things, such as reading the newspaper or watching television",
           "Moving or speaking so slowly that other people could have noticed? Or the opposite — being so fidgety or restless that you have been moving .around a lot more than usual",
           "Thoughts that you would be better off dead or of hurting yourself in some way"),
         ~ factor(as.character(.x), 
                  levels = c("0", "1", "2", "3"),
                  labels = c("Not at all", "Several days", 
                             "More than half days", "Nearly everyday"))))


post_bariatric_depression_data %>% select(
  `Little interest or pleasure in doing things`,
  `Feeling down, depressed, or hopeless`, 
  `Trouble falling or staying asleep, or sleeping too much`, 
  `Feeling tired or having little energy`, 
  `Poor appetite or overeating`, 
  `Feeling bad about yourself — or that you are a failure or have let yourself or your family down`,
  `Trouble concentrating on things, such as reading the newspaper or watching television`,
  `Moving or speaking so slowly that other people could have noticed? Or the opposite — being so fidgety or restless that you have been moving .around a lot more than usual`,
  `Thoughts that you would be better off dead or of hurting yourself in some way`,
  `Score of PHQ-9`
) %>% tbl_summary() %>% bold_labels() %>% as_flex_table() %>% 
  set_table_properties(layout = "autofit", width = 1) %>%
  autofit()
```

  Unadjusted and adjusted logistic regression analyses were conducted to identify predictors of depression following bariatric surgery. No statistically significant predictor was identified in either the unadjusted or adjusted models, as all 95% confidence intervals crossed the null value of 1.0. 

### Table 3: Logistic regression analysis: Factors associated with development of depression after bariatric surgey
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Run separate univariate binary logistic regressions for each predictor
# against the Depression outcome. Results are exponentiated to give
# crude odds ratios (OR) with 95% CIs.
univariate_regressoin <- post_bariatric_depression_data %>% select(
  Age, Gender, `Maraital Status`, `Educational Level`, Hypertension, Diabetes, `Heart Disease`,
  Asthma, `Other Comorbidities`, `High Cholesterol`, `BMI PreOp`, Surgery_Type, `Date since operation`, `Post Surgical Complications`,
  Depression, `Post Surgical Complications`, `Percentage of Weight Loss`
) %>% tbl_uvregression(label = list(
  Surgery_Type ~ "Type of Surgery"
),
method = glm,
  y = Depression, 
  method.args = list(family = binomial),
  exponentiate = T, hide_n = T
) 

# Fit a full multivariable binary logistic regression model including all
# candidate predictors. This produces adjusted odds ratios (aOR).
multivariable_model <- glm(Depression ~ Age + Gender + `Maraital Status`+ 
                             `Educational Level`+ Hypertension+ Diabetes+ 
                             `Heart Disease`+ Asthma + `Other Comorbidities` + 
                             `High Cholesterol`+ `BMI PreOp`+ Surgery_Type + 
                             `Date since operation`+ `Post Surgical Complications`+ `Post Surgical Complications`+ `Percentage of Weight Loss`,
                           data = post_bariatric_depression_data, 
    family = binomial)

# Format the multivariable model output as a gtsummary table.
# p-values <= 0.05 are bolded for quick identification of significant predictors.
multivariate_regression <- tbl_regression(multivariable_model, 
               exponentiate = T) %>% bold_p(t = 0.05)

# Side-by-side display of crude and adjusted ORs for all predictors.
# tab_spanner labels clearly distinguish the two models.
tbl_merge(
  tbls = list(univariate_regressoin, multivariate_regression),
  tab_spanner = c("**Unadjusted**", "**Adjusted**")
) %>% bold_labels() %>% as_flex_table()
```


 
 
 
 